Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-07-14 paper/manuscript.tex
added some more detail to subsampling.

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.

@@

MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ where the sampled distribution is the \textit{mixture distribution} of all the other samples thrown together in one ``pot'', defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the weighted average of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.

@@

@@

\begin{checklist}{Running absolute simulations}
        \textbf{Steps to running your absolute alchemical free energy calculations}
\begin{itemize}
 \item Check your ligands have the same, biologically correct binding pose
        \item Make sure your \textlambda-scheduling is  appropriate
        \item Check if your ligands are discharging and decoupling correctly
        \item Set up your restraints correctly
        \item Make sure you subsample the data in your free energy estimation protocol
        \item Apply the appropriate correction terms
\end{itemize}
\end{checklist}

\begin{checklist}{Running relative simulations}
        \textbf{Steps to running your relative alchemical free energy calculations}
\begin{itemize}
   \item Check your ligands  have the same, biologically correct binding pose
        \item Make sure your $\lambda$-scheduling is set correctly
        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results
        \item Check whether dummy atoms were assigned correctly
        \item Consider subsampling the data in your free energy estimation protocol
        \item Apply the 
        appropriate correction terms
\end{itemize}
\end{checklist}
\end{Checklists*}

@@

\begin{checklist}{Data Analysis}
    \textbf{Steps to analyzing your output data correctly}
\begin{itemize}
    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
    \item Include error bars in all your visual analyses
\end{itemize}
\end{checklist}
\end{Checklists*}
\clearpage

>>>>

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty.

A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.

@@

MBAR can also be thought of as the Zwanzig estimator of the free energy to state $i$ from a sampled distribution referred to as the \textit{mixture distribution}, where one throws all the samples from all simulations together in one ``pot''. Mathematically, this is defined by $p_m(\vec{q}) = N^{-1} \sum_k N_k \exp(f_k-u_k\vec{q})$, which is the average, weighted by the number of samples drawn from each distribution, of all the individual normalized probability distributions from the simulations that are performed~\cite{shirts2017reweighting}.

All of the estimators above have uncertainty estimates associated with them.  In the limit of sufficiently large data, perhaps several hundred or thousand independent points, exhaustive analyses of model data has shown that the error estimates of BAR, MBAR, and TI are all quite reliable when there is good overlap between intermediate states. For EXP, the number of samples at which they become reliable is significantly higher, especially if overlap between intermediate states is lower. However, in the limit of moderate or small amounts of data (tens or hundreds of independent samples) these errors become somewhat unreliable, often off by a factor of 2 from what would get if more data was collected. These estimator of uncertainty are all derived from propagation of error formulae from the estimators themselves. However, when the underlying data is noisy, this approximation is not very good.

Some of the authors' experience has shown that significantly more reliable estimators at lower numbers of samples are obtained by performing bootstrap sampling on data that has been uncorrelated. We do not provide a full tutorial of bootstrapping here, but the basic idea is that one takes the $N_{sample}$ data points collected and draws, with replacement, $N_{sample}$ samples.  This means some of the data points will be represented multiple times, and other data points none. This process is then repeated $N_{bootstrap}$ times, where $N$ should generally be at least 50 to up to 200 for good estimates of the standard error of the estimator.  The data is then \textit{treated} as $N_{bootstrap}$ samples from the true distribution, and statistics such as the standard error are computed from this artificial set of $N_{bootstrap}$ experiments.  In the limit of $N_{sample}$ large, it can be shown that the bootstrap estimate of the standard error converges to the true population standard error.  One way this can be seen is that bootstrap estimators of the data are independent samples of the \textit{actual} distribution of the experiment performed, i.e., a bunch of $\delta$ functions at $N_{sample}$ points, rather than being independent samples from the true, usually continuous distribution.  As more and more data is collected, our collection of $\delta$ functions becomes more and more like the true distribution. Another advantage of bootstrap sampling is that one can construct decent estimates of error bounds with arbitrary percentiles, and not just standard errors of the estimates, which assume Gaussian behavior of the underlying error, and that the error results are Gaussian as well.

One important point is that the bootstrap samples use to estimate the standard error should be decorrelated, or else the bootstrap estimate of the standard error will be artificially small.  More discussion of subsampling and decorrelation is included in section~\ref{sec:decorrelating-samples}. Another alternative to bootstrap sampling of individual points that better captures the correlations in the data is block bootstrapping~\cite{???}.  Block bootstrapping is somewhat more complicated and less common to perform and analyze than bootstrap estimates of the error, so we will not go into details, but it can also provide more reliable estimates of uncertainty than the estimators above.

@@

%Cooment: Finlay, can you add description/citation of RED?

@@

\begin{checklist}{Running absolute simulations}
        \textbf{Steps to running your absolute alchemical free energy calculations}
\begin{itemize}
 \item Check your ligands have the same, biologically correct binding pose
        \item Make sure your \textlambda-scheduling is  appropriate
        \item Check if your ligands are discharging and decoupling correctly
        \item Set up your restraints correctly
        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
        \item Apply the appropriate correction terms
\end{itemize}
\end{checklist}

\begin{checklist}{Running relative simulations}
        \textbf{Steps to running your relative alchemical free energy calculations}
\begin{itemize}
   \item Check your ligands  have the same, biologically correct binding pose
        \item Make sure your $\lambda$-scheduling is set correctly
        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results
        \item Check whether dummy atoms were assigned correctly
        \item Subsample the data to estimate the variance of your free energy estimation protocol (but not to estimate the values themselves)
         \item Apply the appropriate correction terms
\end{itemize}
\end{checklist}
\end{Checklists*}

@@

\begin{checklist}{Data Analysis}
    \textbf{Steps to analyzing your output data correctly}
\begin{itemize}
    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
    \item Include error bars in all your visual analyses
    \item Consider bootstrapping uncertainties rather than using free energy error estimators
\end{itemize}
\end{checklist}
\end{Checklists*}
\clearpage
